A comparison between the latest Prophet forecasting model and the ARIMA model: An evaluation of BTC/ZAR predictability

### 2017-11-06 Draft/Write-up/Draft.Rmd
Final Paper

---
title: "A comparison between the latest Prophet forecasting model and the ARIMA model: An evaluation of BTC/ZAR predictability"
Author: "Fiona Ganie"  
Student: "GNXFIO001"
bibliography: Tex/ref.bib       
output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 4 
    fig_height: 3
abstract: |
  The Autoregressive Integrated Moving Average (ARIMA) model is one of the most widely used time series models that has gained popularity in the exchange rate market due to its ease of implementation and tractability. With the evolution of computational power, soft computing techniques have since been used in exchange rate markets. The ARIMA model has been used as the standard benchmark model against which these complex methods have been compared. Although they have produced significantly better results than the ARIMA model, these models lack interpretability and building them is a challenging task. Prophet is a sophisticated model that differs from traditional time series models in that it can be utilised by non-experts who have little knowledge about the statistical intricacies involved in the model, however have domain knowledge about the data generating process. Prophet is also robust to outliers and can handle missing values in the time series without the need for interpolation. This paper compares the out-of-sample forecasts of Prophet and ARIMA over varying forecast horizons, using a Mincer-Zarnowitz approach for forecast evaluation. A complete dataset of the prices of Bitcoin/Rand as well as a dataset which has missing prices and outliers present is used in the analysis. It is found that the ARIMA model produces the most efficient and unbiased forecasts over all forecast horizons, and that the ARIMA model is more robust when faced with missing values and outliers present in the data. 
---

@@

# Introduction \label{Introduction}
The Autoregressive Integrated Moving Average (ARIMA) model is one of the most widely used time series models that have attracted attention in financial market forecasting [@khashei2009]. Although the Random Walk model has typically been applied to foreign exchange markets and has produced superior results, some researchers have contended that foreign exchange markets are not efficient and believe that future prices depend on current and past events [@abu1996]. This has led to the application of the ARIMA model to exchange rate problems, where it has since gained popularity due to its ease of implementation and tractability. The ARIMA model is easy to use and yields good forecasts over short forecast horizons, however the linearity of the ARIMA model fails to adequately capture the non-linearity inherent in exchange rate data [@zhang1998].

@@

In December 2016, Facebook open-sourced their forecasting model Prophet. Prophet differs from traditional time series model such as the ARIMA in that it can produce high quality forecasts in a straightforward way. Prophet is a sophisticated model that provides informative results, however is configurable and easy to use.  It can be utilised by non-experts who have little knowledge about the statistical intricacies involved in the model, however have domain knowledge about the data generating process. This knowledge can easily be incorporated into the model through its intuitively adjustable parameters [@taylor2017]. Furthermore, Prophet is robust to outliers and can handle missing values in the time series without the need for interpolation [@taylor2017]. This allows analysts without knowledge on how to pre-process the data, to utilise the model without sacrificing predictive accuracy.  If Prophet produces forecasts that are as good as the ARIMA model, it can be compared to more complex forecasting methods that are less tractable and flexible.

@@

## Other techniques used to forecast exchange rates
Generalised autoregressive conditional heteroskedasticity (GARCH) models were later developed due to the failure of ARIMA models to capture the volatility in financial markets  [@anastasakis2009]. @fahimifard2009 applied a GARCH model to the Rial/USD and Rial/EUR to compare the forecasts yielded by the GARCH and ARIMA model over varying forecast horizons.  Their results illustrated that the GARCH model outperformed the ARIMA model when using the Root Mean Square Error (RMSE), Mean Square Error (MSE) and Mean Absolute Deviation (MAD) as a performance criteria.

@@

## Model Confidence Set
The Model Confidence Set (MCS) procedure developed by @hansen2011 is a sequential procedure which starts with an initial set of models $M_0$ of size m. The worst model is then eliminated at each step until the null hypothesis of equal predictive ability (EPA) fails to be rejected for all the models belonging to the set at a given confidence level $1 - \alpha$. This smaller set of models $M^*$ is the superior set of models (SSM) with the best case occurring when the SSM consists of a single best model.

@@

$$t_i,_j = \frac{\bar{d_i,_j}}{\sqrt{\hat{var[\bar{d_i,_j}]}}}$$ \vspace{1pt}
$$ T_R = \max_{i,j\in M} \;\; |t_i,_j|$$
where $\bar{d_i,_j} = m^{-1}\sum_{t=1}^{m}{d_i,_j,_t }$ measures the sample average loss differential between model $i$ and $j$ and  $\hat{var[\bar{d_i,_j}]}$ is estimated through bootstrapping as described in @hansen2011.

This paper uses a block–bootstrap procedure of 5000 resamples as in @hansen2011. The block length $p$ is chosen by fitting an AR(p) model to the $d_i,_j,_t$ terms and determining the maximum number of significant AR parameters obtained. While $t_i,_j$ has a t-distribution, $T_R$ has a non-standard distribution under the null hypothesis hence is estimated by implementing a bootstrapping procedure similar to that used to estimate $\hat{var[\bar{d_i,_j}]}$.

The worst model is then eliminated according to an elimination rule which is coherent with the calculated t-statistic and is defined as:

$$e_R = \arg max_{i}\bigg(\sup_{j\in M}\frac{\bar{d_i,_j}}{\sqrt{\hat{var[\bar{d_i,_j}]}}}\bigg)$$

## Mincer-Zarnowitz Approach to Forecast Evaluation 
The Mincer-Zarnowitz approach to evaluating forecast accuracy is commonly used and can be useful when comparing the forecasts produced by Prophet and the ARIMA model. @mincer1969 proposed an absolute measure to evaluate forecast accuracy that considers the distance between the actual and predicted values. To analyse the absolute errors produced by the forecasts, the observed values $r_t$ are regressed against the predicted values $\hat{r_t}$, i.e. 
$$r_t = \alpha + \beta\cdot\hat{r_t} + e_t$$
where $\alpha$ represents the mean distance between the observed and predicted values while $\beta$ represents the correlation between the forecasted errors and predicted values. When $\alpha = 0$ it implies that the forecasts are unbiased and do not systematically overestimate or underestimate the data. When $\beta = 1$ it implies that the forecasts are efficient and uncorrelated with the forecasted errors. A joint hypothesis test of $H_0: \alpha = 0 \cup \beta = 1$ is performed to check the efficiency and bias of the forecasts and the model which produces the best results generates superior forecasts.

>>>>

---
title: "A comparison between the latest Prophet forecasting model and the ARIMA model: An evaluation of BTC/ZAR predictability"
Author: "Fiona Ganie"  
Student: "GNXFIO001"
bibliography: Tex/ref.bib       
output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 4 
    fig_height: 3
abstract: |
  The Autoregressive Integrated Moving Average (ARIMA) model is one of the most widely used time series models that has gained popularity in the exchange rate market due to its ease of implementation and tractability. With the evolution of computational power, soft computing techniques have since been used in exchange rate markets. The ARIMA model has been used as the standard benchmark model against which these complex methods have been compared. Although they have produced significantly better results than the ARIMA model, these models lack interpretability and building them is a challenging task. Prophet is a sophisticated model that differs from traditional time series models in that it can be utilised by non-experts who have little knowledge about the statistical intricacies involved in the model, however have domain knowledge about the data generating process. Prophet is also robust to outliers and can handle missing values in the time series without the need for interpolation. This paper compares the out-of-sample forecasts of Prophet and ARIMA over varying forecast horizons, using the Model Confidence Set and Mincer-Zarnowitz test to evaluate forecast accuracy. A complete dataset of the prices of Bitcoin/Rand as well as a dataset which has missing prices and outliers present is used in the analysis. It is found that the ARIMA model ranks ahead of Prophet over all forecast horizons and is more robust when faced with missing values and outliers. In addition, both models having equal predictive accuracy when forecasting at short and long forecast horizons when the data is clean and at short forecast horizons when the data is dirty, with unbiased and/or efficient forecasts only generated over the longest forecast horizon in the case of the clean data. 
---

@@

# Introduction \label{Introduction}
The Autoregressive Integrated Moving Average (ARIMA) model is one of the most widely used time series models that has attracted attention in financial market forecasting [@khashei2009]. Although the Random Walk model has typically been applied to foreign exchange markets and has produced superior results, some researchers have contended that foreign exchange markets are not efficient and believe that future prices depend on current and past events [@abu1996]. This has led to the application of the ARIMA model to exchange rate problems, where it has since gained popularity due to its ease of implementation and tractability. The ARIMA model is easy to use and yields good forecasts over short forecast horizons, however the linearity of the ARIMA model fails to adequately capture the non-linearity inherent in exchange rate data [@zhang1998].

@@

In December 2016, Facebook open-sourced their forecasting model Prophet. Prophet differs from traditional time series models such as the ARIMA in that it can produce high quality forecasts in a straightforward way. Prophet is a sophisticated model that provides informative results, however is configurable and easy to use.  It can be utilised by non-experts who have little knowledge about the statistical intricacies involved in the model, however have domain knowledge about the data generating process. This knowledge can easily be incorporated into the model through its intuitively adjustable parameters [@taylor2017]. Furthermore, Prophet is robust to outliers and can handle missing values in the time series without the need for interpolation [@taylor2017]. This allows analysts without knowledge on how to pre-process the data, to utilise the model without sacrificing predictive accuracy.  If Prophet produces forecasts that are as good as the ARIMA model, it can be compared to more complex forecasting methods that are less tractable and flexible.

@@

## Other techniques used to forecast exchange rates
Generalised autoregressive conditional heteroskedasticity (GARCH) models were later developed due to the failure of ARIMA models to capture the volatility in financial markets  [@anastasakis2009]. @fahimifard2009 applied a GARCH model to the Rial/USD and Rial/EUR to compare the forecasts yielded by the GARCH and ARIMA model over varying forecast horizons.  Their results illustrated that the GARCH model outperformed the ARIMA model when using the Root Mean Square Error (RMSE), Mean Square Error (MSE) and Mean Absolute Deviation (MAD) as performance criteria.

@@

## Model Confidence Set
The Model Confidence Set (MCS) for a set of models is analogous to a confidence interval for a model parameter. The MCS procedure developed by @hansen2011 is a sequential procedure that starts with an initial set of models $M_0$ of size m. The worst model is then eliminated at each step until the null hypothesis of equal predictive ability (EPA) fails to be rejected for all the models belonging to the set at a given confidence level $1 - \alpha$. This smaller set of models $M^*$ is the superior set of models (SSM) with the best case occurring when the SSM consists of a single best model.

@@

$$t_i,_j = \frac{\bar{d_i,_j}}{\sqrt{\hat{var[\bar{d_i,_j}]}}}$$ \vspace{1pt}
$$ T_R = \max_{i,j\in M} \;\; |t_i,_j|$$
where $\bar{d_i,_j} = m^{-1}\sum_{t=1}^{m}{d_i,_j,_t }$ measures the sample average loss differential between model $i$ and $j$ and  $\hat{var[\bar{d_i,_j}]}$ is estimated through the bootstrapping procedure described in @hansen2011.

This paper uses a block-bootstrap procedure of 5000 resamples. The block length $p$ is chosen by fitting an AR(p) model to the $d_i,_j,_t$ terms and determining the maximum number of significant AR parameters obtained. While $t_i,_j$ has a t-distribution, $T_R$ has a non-standard asymptotic distribution under the null hypothesis hence is estimated by implementing a bootstrapping procedure similar to that used to estimate ${var[\bar{d_i,_j}]}$.

The worst model is then eliminated according to an elimination rule that is coherent with the calculated t-statistic and is defined as:

$$e_R = \arg max_{i}\bigg(\sup_{j\in M}\frac{\bar{d_i,_j}}{\sqrt{\hat{var[\bar{d_i,_j}]}}}\bigg)$$
## Mincer-Zarnowitz Approach to Forecast Evaluation 
The Mincer-Zarnowitz approach to evaluating forecast accuracy is commonly used and can be useful when comparing the forecasts produced by Prophet and the ARIMA model. @mincer1969 proposed an absolute measure to evaluate forecast accuracy that considers the distance between the actual and predicted values. To analyse the absolute errors produced by the forecasts, the observed values $r_t$ are regressed against the predicted values $\hat{r_t}$, i.e. 
$$r_t = \alpha + \beta\cdot\hat{r_t} + e_t$$
where $\alpha$ represents the mean distance between the observed and predicted values while $\beta$ represents the correlation between the forecasted errors and predicted values. When $\alpha = 0$ it implies that the forecasts are unbiased and do not systematically overestimate or underestimate the data. When $\beta = 1$ it implies that the forecasts are efficient and uncorrelated with the forecasted errors. A joint hypothesis test of $H_0: \alpha = 0 \cup \beta = 1$ is performed to check the efficiency and bias of the forecasts and the model which produces the best results generates superior forecasts.
